---
title: "Course Project 1"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and Preprocessing Data
```{r data}
library(dplyr)
library(ggplot2)
activity <- read.csv("activity.csv")
activity$date <- as.Date(activity$date)
```


## Question 1: Mean Steps per Day
```{r q1}
#calculate total
total <- sum(activity$steps, na.rm = T) 

#plot histogram of daily total
hold <- na.omit(activity)
temp <- group_by(hold, date)
dayTot <- summarize(temp, Total = sum(steps, na.rm = T))
ggplot(dayTot, aes(Total)) + geom_histogram() + ggtitle("Histogram of Total Steps per Day")

#calculate mean daily total and median daily total
meanSteps <- mean(dayTot$Total, na.rm = T)
medSteps <- median(dayTot$Total, na.rm = T)
```
The overall total number of steps is `r total`.

The mean total number of steps taken per day is `r meanSteps` steps and the median total number of steps taken per day is `r medSteps` steps.

## Question 2: Average Daily Activity Pattern
```{r q2}
temp <- group_by(hold, interval)
dailyAct <- summarize(temp, Average = mean(steps, na.rm = T))
ggplot(dailyAct, aes(y = Average, x = interval)) + geom_line() + 
    xlab("5-Min Interval") + ggtitle("Average Daily Activity Pattern") +
    ylab("Average Number of Steps")

maxInt <- as.integer(dailyAct[which.max(dailyAct$Average),"interval"])
```
The interval `r maxInt` contains the maximum number of steps on average across all the days in the dataset.

## Question 3: Imputing Missing Values
```{r q3}
nas <- filter(activity, is.na(steps))
numRows <- nrow(nas)

##strategy will be to use the average of a given interval to fill in the missing value
temp <- group_by(activity, interval)
dailyAct <- summarize(temp, Average = mean(steps, na.rm = T))

#fill in missing values
for (i in unique(nas$interval)) nas[nas$interval == i,"steps"] = dailyAct[dailyAct$interval == i, "Average"]

#remove missing values from the base dataset, and then re-add the rows with the new values
activity <- na.omit(activity)
activity <- rbind(activity, nas)

#calculate total steps
newTotal <- sum(activity$steps, na.rm = T) 

#plot histogram of daily total
temp <- group_by(activity, date)
dayTot <- summarize(temp, Total = sum(steps, na.rm = T))
ggplot(dayTot, aes(Total)) + geom_histogram() + ggtitle("Histogram of Total Steps per Day")

#calculate mean daily total and median daily total
newMeanSteps <- mean(dayTot$Total, na.rm = T)
newMedSteps <- median(dayTot$Total, na.rm = T)

dif = newTotal - total
```
The number of rows with missing values is `r numRows`.

The new total number of steps is `r newTotal` which is `r dif` more steps than the old total number of steps, `r total`. 

Also, the new median number of steps per day is `r newMedSteps` and the new mean is `r newMeanSteps`. The mean is the same while the median is slightly higher.


```{r q4}
## Create a factor variable that splits dates into weekend and weekday
activity$dayType <- "Weekend"
check <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

activity[weekdays(activity$date) %in% check,"dayType"] <- "Weekday"

activity$dayType <- as.factor(activity$dayType)

#plot Average Daily Activity Pattern differentiated by weekend vs. weekday
temp <- group_by(activity, interval, dayType)
weekCheck <- summarize(temp, Average = mean(steps, na.rm = T))
ggplot(weekCheck, aes(y = Average, x = interval)) + geom_line() + 
    xlab("5-Min Interval") + 
    ggtitle("Average Daily Activity Pattern for Weekends vs. Weekdays") +
    facet_grid(dayType~.) + ylab("Average Number of Steps")
```